Title Goes Here

### 2021-02-10 analysis/paper/paper.Rmd
Tabla

library(metR)
library(data.table)
library(magrittr)
library(ggplot2)
library(tagger)
library(ggperiodic)
library(patchwork)
library(shceof)

@@

If

@@

@@

@@

(ref:cap-splitted) Coefficient of determination of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level. The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time. (probably not to be shown.)

```{r corr-ceof-splitted, fig.cap = "(ref:cap-splitted)"}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag := abs(hgt)] %>%
  .[, item := paste(lev, PC, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  ggplot(aes(item1, item2)) +
  geom_raster(aes(fill = correlation^2)) +
  shadowtext::geom_shadowtext(aes(label = round(correlation^2, 2)),
                              color = "black", bg.colour = "white", bg.r = 0.2) +
  scale_fill_gradient(guide = "none", low = "white", high = scales::muted("red")) +
  scale_x_discrete("200 hPa") +
  scale_y_discrete("50 hPa")
```

@@

```{r mean-anom, fig.cap = "Mean SON zonal geopotential height anomalies at (a) 50 hPa and (b) 200 hPa"}
era5 %>% 
  .[, mean(-hgt), by = .(lon, lat, lev)] %>% 
  .[lev %in% c(50, 200)] %>% 
  .[lat <= -20] %>% 
  .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>% 
  periodic(lon = c(0, 360)) %>% 
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
  geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
  geom_qmap(~ .x[lat <= -20]) +
  scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0), 
                       guide = guide_colorstrip_bottom()) +
  scale_x_longitude() +
  scale_y_latitude() +
  facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
  coord_polar() +
  axis_labs_smol +
  tag_facets()
```

@@

@@

The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@(ref:fig-extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`. However, there is no trend in the hybrid dataset during the period of overlap (p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset != "era20c"]$p_report`) and the correlations between the old and modern series are much poorer for this variable. This suggests that the observed trend is not robust.

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2020a. Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

@@

sam_r2 <- sams %>% 
  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, with(cor.test(hgt, estimate), 
           list(estimate = estimate, 
                low = conf.int[1],
                high = conf.int[2],
                p_value = p.value)), by = .(lev, term, part, PC)] %>%

  .[, low2 := pmin(low^2, high^2, ifelse(sign(high) != sign(low), 0, 1))] %>% 
  # .[low2 <= 0, low2 := 0] %>% 
  .[, high2 := pmax(low^2, high^2)] %>% 
  .[PC != "PC3"]

@@

sam_r2 %>% 
  ggplot(aes(lev, estimate^2)) +
  # geom_ribbon(data =  ~.x[p.adjust(p_value, "fdr") < 0.001], 
  # aes(ymin = low2, ymax = high2, color = term), alpha= 0.1) +
  geom_point(data = ~.x[p.adjust(p_value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²", limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc")
```

Figure \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices. The SAM index represents the leading EOF of monthly geopotential height fields south of 20ºS at each level. The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2020a). Since the analysis here is only for the SON trimester, monthly values were averaged across semesters (weighted by the number of days in each month).

@@

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(data = sst_gdata, 
                        aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour2(data = sst_gdata, 
                  aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(data = sst_gdata,
                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 3)),
                geom = "point", size = 0.2, alpha = 0.5) +
    scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ var, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
    coord_quickmap(ylim = c(NA, 10)) +
    theme(legend.box = "horizontal") +
    # ggh4x::facet_nested(term ~ PC) +
    axis_labs_smol +
    tag_facets("cr")
}
```

```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_sst_psi(which_pc = "PC2", which_lev = 200) 
# +   annotate("rect", xmin = 170, xmax = 195, ymin = -25, ymax = -5, color = "black", fill = NA)
```

??? -\> esta intro hay que cambiarla. To help identify those sources Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and STT and Streamfunction at 200 hPa. As expected from the correlations between each component of EOF2 and ENSO, Imaginary EOF2 is associated with strong ENSO-like SST anomalies and streamfunction (Figure \@ref(fig:sst-psi-2).a2). Streamfunction anomalies (Figure \@ref(fig:sst-psi-2).b2) are indicative of tropical forcing, which is compatible with SST anomalies. Real EOF2, on the other hand, is not associated with significant SST anomalies (Figure \@ref(fig:sst-psi-2).a1), as expected by its lack of relationship with ENSO, and streamfunction anomalies in the tropics are much weaker and non significant.

```{r enso_cor}
enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
    sep_ReIm() %>%
    .[, hgt := scale(hgt), by = .(PC, part)] %>%
    enso[., on = "time"] %>%
    na.omit() %>% 
   .[, with(cor.test(hgt, oni), list(estimate = estimate,
                                     p.value = p.value,
                                     low = conf.int[1],
                                     hig = conf.int[2])), 
     by = .(PC, part)] %>% 
  .[, cor_text := paste0(signif(estimate, 2), " (CI: ", 
                         signif(low, 2), " -- ", signif(hig, 2), ")")]

```{r enso-phase, fig.cap = "Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019. Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile. Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2."}
ceof[, eof[[1]]$left, by = .(season)] %>% 
  .[, arg := Arg(hgt)] %>% 
  .[, mag := Mod(hgt)] %>% 
  .[enso, on = "time"] %>% 
  na.omit() %>% 
  .[PC == "PC2"] %>% 
  .[year(time) >= 1979] %>%
  qwrap(arg = c(-pi, pi) ~ c(-2*pi, 2*pi)) %>% 
  ggplot(aes(arg, oni)) +
  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
  geom_smooth(method = "lm", aes(weight = mag),
              formula = y ~ I(cos(x)) + I(sin(x)), 
              se = FALSE, color = "black", size = 0.5) +
  geom_point(aes(colour = mag > quantile(mag, 0.5)), size = 3) +
  scale_color_brewer(NULL, palette = "Dark2", 
                     labels = c("TRUE" = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_y_continuous("ONI") +
  scale_x_continuous("Phase", 
                     breaks = seq(-pi, pi, by = 30*pi/180), 
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  coord_cartesian(xlim = c(-pi, pi))
```

The zonal wave represented by the EOF2 pattern is very similar to the wave train associated with ENSO. Indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is `r enso_cor[PC == "PC2" & part == "Imaginary"]$cor_text`. The correlation between the Real EOF2 and ENSO is not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`). This suggests that, even though both EOF2 patterns represent the same wave train with a different phase, these two phases are exited by different sources.

Figure \@ref(fig:sst-psi-2) suggests, then, that the wave train represented by EOF2 can be forced both by the tropics and internally, and that it has preferred phases in each case. When it is forced by the tropics, it aligns roughly with the Imaginary phase, and when it's the Real phase is closer to the internally-varying wave. However, Figure \@ref(fig:sst-psi-2).b1 also indicates that this separation is not completely clean. It's not expected.. ??

@@

>>>>

library(metR)
library(data.table)
library(magrittr)
library(ggplot2)
library(tagger)
library(ggperiodic)
library(patchwork)
library(shceof)
library(kableExtra)

@@

@@

No mostrar los EOFs no complejos. No tiene sentido mostralos si no se analizas.

Las correlaciones entre PCs Podrían ir en una tabla. Quizás suplementaria.

Relacionar 50hPa con vórtice polar? Temperatura también suma. 
En 50 el negativo es más intenso que el negativo. ¿Por qué?

Ver qué pasa con U con el EOF1.

En 200 no se ve la señal de jet de la temperatura ¿qué pasa más abajo? 
en 700 la anomalía de temp coincide más o menos con el geopotentical

ondas antes e la tempreatura

Mostrar EOF espaciales.

Tirar ERA20C a la basura. ERA20c suplementaria?

Pp, ¿cuál es la regresión con SAM asim en 200 hPa?

@@

```{r}

background_palette <- grDevices::colorRampPalette(c("white", scales::muted("red")), 
                                                  space = "Lab")

eof_naive[, eof[[1]]$left, by = lev] %>% 
  .[, item := paste(lev, EOF, sep = "_")] %>% 
  widyr::pairwise_cor(item, time, hgt) %>% 
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>% 
  .[tstrsplit(item1, "_")[[1]] != 50] %>% 
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>% 
  .[, r2 := round(correlation^2, 2)] %>% 
  dcast(item1 ~ item2, value.var = "r2") %>% 
  setnames("item1", "200 hPa") %>% 
  kbl2(booktabs = TRUE, caption = "$R^2$ between pairs of EOFs of each level rounded to two decimals.") %>% 
  # add_header_above(c("", "50 hPa" = 4)) %>% 
  columns_spec(2:5,  background = spec_color(.table[[.col]], option = background_palette, 
                                              scale_from = c(0, 1)),
               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
  add_header_above(c("", "50 hPa" = 4)) %>%
  kable_classic()
```

@@

```{r corr-ceof-splitted}
ceof_splitted[, eof[[1]]$left, by = .(lev = levs)] %>%
  copy() %>%
  .[, mag := abs(hgt)] %>%
  .[, item := paste(lev, PC, sep = "_")] %>%
  widyr::pairwise_cor(item, time, mag) %>%
  setDT() %>%
  # .[]
  .[tstrsplit(item1, "_")[[1]] != tstrsplit(item2, "_")[[1]], ] %>%
  .[tstrsplit(item1, "_")[[1]] != 50] %>%
  .[, item1 := tstrsplit(item1, "_")[[2]]] %>%
  .[, item2 := tstrsplit(item2, "_")[[2]]] %>%
  .[, r2 := round(correlation^2, 2)] %>%
  dcast(item1 ~ item2, value.var = "r2") %>%
  setnames("item1", "200 hPa") %>%
  kbl2(booktabs = TRUE, caption = "$R^2$ of the absolute magnitude of complex EOFs between 200 hPa and 50 hPa computing EOF separatedly for each level. The high correlation between levels for the two leading EOFs justifies treating the pattern as a mode of covariability between the stratosphere and the troposphere and, thus, computing the EOFs using both levels at the same time.") %>%
  columns_spec(2:4, background = spec_color(.table[[.col]], option = background_palette,
                                            scale_from = c(0, 1)),
               color = ifelse(.table[[.col]] > .5, "white", "black")) %>% 
  add_header_above(c("", "50 hPa" = 3)) %>%
  kable_classic() 
```

@@

```{r mean-anom, fig.cap = "Mean SON zonal geopotential height anomalies at (a) 50 hPa and (b) 200 hPa"}
# era5 %>%
#   .[, mean(-hgt), by = .(lon, lat, lev)] %>%
#   .[lev %in% c(50, 200)] %>%
#   .[lat <= -20] %>%
#   .[, V2 := V1 - mean(V1), by = .(lat, lev)] %>%
#   periodic(lon = c(0, 360)) %>%
#   ggplot(aes(lon, lat)) +
#   geom_contour_fill(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
#   geom_contour_tanaka(aes(z = V2), global.breaks = FALSE, breaks = AnchorBreaks(exclude = 0)) +
#   geom_qmap(~ .x[lat <= -20]) +
#   scale_fill_divergent(NULL, breaks = AnchorBreaks(exclude = 0),
#                        guide = guide_colorstrip_bottom()) +
#   scale_x_longitude() +
#   scale_y_latitude() +
#   facet_grid(~lev, labeller = labeller(lev = lab_lev)) +
#   coord_polar() +
#   axis_labs_smol +
#   tag_facets()
```

@@

```{r}

@@

The ERA20C time-series has a also significant upward linear trend in the imaginary part of EOF1 (Figure \@ref(fig:extended-series)a.2, p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset == "era20c"]$p_report`. However, there is no trend in the hybrid dataset during the period of overlap (p-value `r trends[part == "Imaginary" & PC == "PC1" & dataset != "era20c"]$p_report`) and the correlations between the old and modern series are much poorer for this variable. This suggests that the observed trend is not robust.

@@

(ref:sam-eof-vertical-cap) Coefficient of determination between the real and imaginary part of each EOF and the SAM, Asymmetric SAM (A-SAM) and Symmetric SAM (S-SAM) indices computed at each level according to @campitelli2021. Points mark estimates with p-value \< 0.01 corrected for False Detection Rate [@benjamini1995].

@@

sam_r2 <- sams %>%
  .[time, on = .NATURAL, allow.cartesian = TRUE] %>%
  .[, correlate(hgt, estimate), by = .(lev, term, part, PC)]

@@

sam_r2 %>%
  ggplot(aes(lev, estimate^2)) +
  # geom_ribbon(data =  ~.x[p.adjust(p_value, "fdr") < 0.001],
  # aes(ymin = low2, ymax = high2, color = term), alpha= 0.1) +
  geom_point(data = ~.x[p.adjust(p.value, "fdr") < 0.01], aes(color = term)) +
  geom_line(aes(color = term, size = term)) +
  scale_x_level() +
  scale_y_continuous("r²", limits = c(0, 1)) +
  scale_size_manual(guide = "none",
                    values = c(full = 1.2,
                               asym = 0.5,
                               sym = 0.5)) +
  scale_color_brewer(NULL, palette = "Dark2", labels = c(full = "SAM",
                                                         asym = "A-SAM",
                                                         sym = "S-SAM")) +
  facet_grid(part ~ PC, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
  coord_flip() +
  tag_facets("rc")
```

Figure \@ref(fig:sam-eof-vertical) shows the coefficient of determination between the EOFs and three SAM indices. The SAM index represents the leading EOF of monthly geopotential height fields south of 20ºS at each level. The A-SAM and S-SAM indices represent, respectively, the zonally asymmetric and symmetric component of the SAM and are obtained by projecting the zonally asymmetric and zonally symmetric part of the SAM spatial pattern onto monthly geopotential height fields (@campitelli2021). Since the analysis here is only for the SON trimester, monthly values were averaged across semesters (weighted by the number of days in each month).

@@

    geom_contour_fill(data = sst_gdata,
                      aes(z = estimate, fill = ..level..), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour_tanaka(data = sst_gdata,
                        aes(z = estimate), breaks = AnchorBreaks(exclude = 0)) +
    geom_contour2(data = sst_gdata,
                  aes(z = p_val), breaks = 0.05, size = 0.1) +
    stat_subset(data = sst_gdata,
                aes(subset = p_val <= 0.05 & is.cross(lon, lat, 3)),
                geom = "point", size = 0.2, alpha = 0.5) +
    scale_fill_divergent_discretised("SST", guide = guide_colorsteps_bottom(order = 2)) +
    scale_x_longitude() +
    scale_y_latitude() +
    scale_mag(guide = "none") +
    facet_grid(term ~ var, labeller = labeller(PC = c(PC1 = "EOF1", PC2 = "EOF2"))) +
    coord_quickmap(ylim = c(NA, 10)) +
    theme(legend.box = "horizontal") +
    # ggh4x::facet_nested(term ~ PC) +
    axis_labs_smol +
    tag_facets("cr", position = "bl")
}
```

```{r enso_cor}
enso_cor <-  ceof[, eof[[1]]$left, by = .(season)] %>%
  sep_ReIm() %>%
  .[, hgt := scale(hgt), by = .(PC, part)] %>%
  enso[., on = "time"] %>%
  na.omit() %>%
  .[, correlate(hgt, oni), by = .(PC, part)]
```

```{r sst-psi-2,  fig.cap = "Regresión con PSI en 200 hPa y sst, PC1"}
plot_sst_psi(which_pc = "PC2", which_lev = 200)
```

??? -\> esta intro hay que cambiarla. Figure \@ref(fig:sst-psi-2) shows regression maps between EOF2 and Sea Surface Temperatures (SST) Streamfunction at 200 hPa. The Imaginary EOF2 is associated with strong negative SST anomalies on the Central Pacific and positive anomalies on an area across the North of Australia and New Zealand, the South Pacific Convergence Zone (SPCZ) (Figure \@ref(fig:sst-psi-2).a2). This pattern is almost canonically negative ENSO and indeed, the correlation between the Imaginary EOF2 and the Oceanic ENSO Index ([@bamston1997]) is `r enso_cor[PC == "PC2" & part == "Imaginary"]$cor_text`. Streamfunction anomalies show a coherent picture. The Imaginary EOF2 is associated with strong wave-like streamfunction anomalies emanating from the tropics (Figure \@ref(fig:sst-psi-2).b2). . This is consistent with what we know of the effect of ENSO on the extratropics: SST anomalies initiate anomalous convection that excites meridionally-moving Rossby waves.

One could expect that the Real EOF2, being just a different phase of the same wave train would show a similar pattern but, perhaps with a slight translation of the main tropical sources. However, the Real EOF2 is associated neither with any significant SST (Figure \@ref(fig:sst-psi-2).a1) nor streamfunction anomalies in the tropics. The correlation between the Real EOF2 and ENSO is also not significant (`r enso_cor[PC == "PC2" & part == "Real"]$cor_text`).

```{r enso-phase, fig.cap = "Relationship between ENSO and phase of EOF2 for the period 1979 -- 2019. Colours denote years with magnitude of EOF2 greater or smaller than the 50th percentile. Black line is the fit ONI ~ cos(phase) + sin(phase) computed by OLS weighted by the magnitude of EOF2."}
cor_enso_mag <- ceof[, eof[[1]]$left, by = .(season)] %>%
  .[, arg := Arg(hgt)] %>%
  .[, mag := Mod(hgt)] %>%
  .[enso, on = "time"] %>%
  na.omit() %>%
  .[PC == "PC2"] %>%
  .[, correlate(mag, abs(oni))]

ceof[, eof[[1]]$left, by = .(season)] %>%
  .[, arg := Arg(hgt)] %>%
  .[, mag := Mod(hgt)] %>%
  .[enso, on = "time"] %>%
  na.omit() %>%
  .[PC == "PC2"] %>%
  ggplot(aes(arg, oni)) +
  geom_hline(yintercept = c(-0.5, 0, 0.5), size = 0.2, color = "gray50") +
  geom_vline(xintercept = 0, size = 0.2, color = "gray50") +
  annotate(shadowtext:::GeomShadowText,
           label = c("- Real", "- Imaginary", "+ Real", "+ Imaginary"),
           color = "gray60", bg.colour = "white", bg.r = 0.2,
           size = 6,
           y = -2, x = c(-pi, -pi/2, 0, pi/2), angle = 90, hjust = 0) +

  geom_smooth(data = ~qwrap(.x, arg = c(-pi, pi) ~ c(-2*pi, 2*pi)),
              method = "lm", aes(weight = mag),
              formula = y ~  I(cos(x)) + I(sin(x)),
              se = FALSE, color = "black", size = 0.5) +
  geom_point(aes(colour = mag >= quantile(mag, 0.5),
                 shape  = mag >= quantile(mag, 0.5)),
             size = 4) +

  scale_color_brewer(NULL, palette = "Dark2",
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_shape_manual(NULL, values = c("TRUE" = 20,
                                      "FALSE" = 18),
                     labels = c("TRUE"  = "|EOF2| > 50%",
                                "FALSE" = "|EOF2| < 50%")) +
  scale_y_continuous("ONI") +
  scale_x_continuous("Phase",
                     breaks = seq(-pi, pi, by = 30*pi/180),
                     labels = function(x) paste0(round(x*180/pi, 5), "º")) +
  coord_cartesian(xlim = c(-pi, pi))
```

We show a more complete picture of the relationship between ENSO and EOF2 in Figure \@ref(fig:enso-phase), which plots the ONI index value and the phase of the EOF2 for each year between 1979 and 2019 highlighting years in which the magnitude of EOF2 was higher than 50% of the years. In years with positive ENSO, the phase of the EOF2 is always around -90º (corresponding with negative imaginary part) and vice versa. In years with near neutral ENSO, the phase of the EOF2 is much more variable between 0º and -180º. There is also a relationship between the magnitude of the EOF2 and the magnitude of the ENSO. The strong EOF2 years are almost invariably associated with strong ENSO years. The correlation between the absolute magnitude of the ONI and the magnitude of the EOF2 is `r cor_enso_mag$cor_text`.

All this suggests that the wave train represented by EOF2 can be both forced by tropical SSTs and also a part of the internal variability of the extratropical atmosphere. When internally forced, there is little phase preference and amplitudes are relatively low. However, this same wave train, when excited by tropical SST anomalies will prefer the imaginary phase, with the sign of the geopotential anomalies depending on the sign of the tropical SSTs anomalies, and can reach higher magnitudes.

```{r  sst-psi-1, fig.cap = "Same as figure befoe but for EOF1"}
plot_sst_psi(which_pc = "PC1", which_lev = 200)
```

```{r}
psi %>%
  .[lev == 200] %>%
  .[, .(psi.z = mean(psi.z)), by = .(lon, lat)] %>%
  .[, c("fx", "fy") := WaveFlux(.SD)] %>%
  .[lat <= 0] %>%
  ggplot(aes(lon, lat)) +
  geom_contour_fill(aes(z = psi.z)) +
  geom_streamline(aes(dx = fx, dy = fy)) +
  geom_qmap(~.x[lat<0], keep = .015) +
  scale_fill_divergent()
```

@@

## Naive EOFs
